# Análise Exploratória dos Dados {#sec-exploratory-data-analysis}

```{r}
#| echo: false
#| results: asis

source("_common.R")
mensagem_capitulo_sem_traducao()
```

## Introdução

Esse capítulo mostrará a você como usar visualização e transformação para explorar seus dados em uma forma sistemática, uma tarefa que os estatísticos chamam de análise exploratória dos dados, ou EDA (Exploratory Data Analysis) abreviamente.
EDA é um ciclo iterativo.
Você:

1.  Gera questões sobre os seus dados.

2.  Encontra respostas visualizando, transformando e modelando seus dados.

3.  Usa o que aprendeu para refinar suas questões e/ou gerar novas questões.

A EDA não é um processo formal com um conjunto estrito de regras.
Mais que qualquer coisa, EDA é um estado da mente.
Durante as fases inicias da EDA você deverá se sentir livre para investigar cada ideia que ocorre a você.
Algumas dessas ideias darão certo e outras serão becos sem saída.
À medida que sua exploração continua, você se concentrará em alguns insights particularmente produtivos que eventualmente escreverá e comunicará a outras pessoas.

A EDA é uma parte importante de qualquer análise de dados, mesmo que as principais questões de pesquisa sejam entregues a você de bandeja, porque você sempre precisa investigar a qualidade dos seus dados.
Limpeza de dados é apenas uma das aplicações da EDA: você se questiona sobre se seus dados vão de encontro às suas expectativas ou não.
Para fazer limpeza de dados, você irá necessitar implementar todas as ferramentas da EDA: visualização, transformação e modelagem.

### Pré-requisitos

Nesse capítulo, nós iremos combinar o que você tem aprendido sobre dplyr e ggplot2 para fazer perguntas interativamente, responder com dados e, em seguida, fazer novas perguntas.

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## Questões

> "Não existem questões estatísticas rotineiras, apenas rotinas estatísticas questionáveis." --- Sir David Cox

> "É muito melhor uma resposta aproximada à pergunta certa, que muitas vezes é vaga, do que uma resposta exata à pergunta errada, que sempre pode ser precisa." --- John Tukey

Seu objetivo durante a EDA é desenvolver uma compreensão dos seus dados.
A maneira mais fácil de fazer isso é usar perguntas como ferramentas para orientar sua investigação.
Quando você faz uma pergunta, ela concentra sua atenção em uma parte específica do seu conjunto de dados e ajuda você a decidir quais gráficos, modelos ou transformações fazer.

EDA é fundamentalmente um processo criativo.
E como a maioria dos processos criativos, a chave para fazer perguntas de *qualidade* é gerar uma grande *quantidade* de perguntas.
É difícil fazer perguntas reveladoras no início da sua análise porque você não sabe quais insights podem ser obtidos do seu conjunto de dados.
Por outro lado, cada nova questão que você fizer irá expor você a um novo aspecto dos seus dados e aumentar sua chance de fazer uma descoberta.
Você pode detalhar rapidamente as partes mais interessantes dos seus dados---e desenvolver um conjunto de perguntas instigantes---se você acompanhar cada pergunta com uma nova pergunta com base no que encontrar.

Não há regra sobre quais perguntas você deve fazer para orientar sua pesquisa.
No entanto, dois tipos de perguntas sempre serão úteis para fazer descobertas em seus dados.
Você pode formular essas perguntas livremente como:

1.  Que tipo de variação ocorre dentro das minhas variáveis?

2.  Que tipo de covariação ocorre entre as minhas variáveis?

O restante deste capítulo examinará essas duas questões.
Explicaremos o que são variação e covariação e mostraremos várias maneiras de responder a cada pergunta.

## Variação

**Variação** é a tendência dos valores de uma variável mudar de medida para medida.
Você pode ver variações facilmente na vida real; se você medir qualquer variável contínua duas vezes, você obterá dois diferentes resultados.
Isso é verdade mesmo se você mede quantidades constantes, como a velocidade da luz.
Cada uma de suas medidas irá incluir uma pequena quantidade de erro que varia de medida a medida.
Variáveis também muda se você mede diferentes objetos (por exemplo, cor dos olhos de diferentes pessoas) ou diferentes tempos (por exemplo, os níveis de energia de um elétron em diferentes momentos).
Cada variável tem seu próprio padrão de variação, que pode revelar interessantes informações sobre como isso varia entre medições na mesma observação, bem como entre observações.
A melhor forma de entender o padrão é visualizando a distribuição dos valores da variável, que você tem aprendido sobre no @sec-data-visualization.

Vamos iniciar nossa exploração visualizando a variação nos pesos (`quilate`) de \~54,000 diamantes do banco de dados `dados`.
Como `quilate` é uma variável numérica, podemos usar um histograma:

```{r}
#| fig-alt: |
#|   A histogram of carats of diamonds, with the x-axis ranging from 0 to 4.5 
#|   and the y-axis ranging from 0 to 30000. The distribution is right skewed 
#|   with very few diamonds in the bin centered at 0, almost 30000 diamonds in 
#|   the bin centered at 0.5, approximately 15000 diamonds in the bin centered 
#|   at 1, and much fewer, approximately 5000 diamonds in the bin centered at 
#|   1.5. Beyond this, there's a trailing tail.

library(dados)

ggplot(diamante, aes(x = quilate)) +
  geom_histogram(binwidth = 0.5)
```

Agora que você pode visualizar a variação, o que você deve observar em seus gráficos?
E que tipo de perguntas de acompanhamento você deve fazer?
Nós temos colocado abaixo uma lista dos tipos mais comuns de informações que você encontrará em seus gráficos, junto com algumas perguntas de acompanhamento para cada tipo de informação.
A chave para fazer boas questões de acompanhamento irá depender da curiosidade (o que mais você quer aprender?) bem como do seu ceticismo (Como isso pode ser enganoso?).

### Valores típicos

Em ambos gráficos de barras e histogramas, barras altas mostram valores comuns de uma variável, e barras mais curtas mostram valores menos comuns.
Locais que não tem barras revelam valores que não foram vistos nos seus dados.
Para transformar essas informações em perguntas úteis, observe qualquer coisa não esperada:

-   Quais valores são os mais comuns?
    Por quê?

-   Quais valores são raros?
    Por quê?
    Isso corresponde às suas expectativas?

-   Você consegue ver padrões incomuns?
    O que pode explicar eles?

Vamos observar a distribuição de `quilate` para os diamantes menores.

```{r}
#| fig-alt: |
#|   A histogram of carats of diamonds, with the x-axis ranging from 0 to 3 and 
#|   the y-axis ranging from 0 to roughly 2500. The binwidth is quite narrow 
#|   (0.01), resulting in a very large number of skinny bars. The distribution 
#|   is right skewed, with many peaks followed by bars in decreasing heights, 
#|   until a sharp increase at the next peak.

menores_diamantes <- diamante |> 
  filter(quilate < 3)

ggplot(menores_diamantes, aes(x = quilate)) +
  geom_histogram(binwidth = 0.01)
```

Esse histograma sugere várias questões interessantes:

-   Por que há mais diamantes em números inteiros de quilates e frações comuns de quilates?

-   Por que há mais diamantes ligeiramente à direita de cada pico do que há ligeiramente à esquerda de cada pico?

Visualizações também podem revelar agrupamentos, os quais sugerem que existem subgrupos em seus dados.
Para entender os subgrupos, questione:

-   Como são as observações dentro de cada subgrupo similares uns aos outros?

-   Como são as observações em grupos diferentes separados uns dos outros?

-   Como você pode explicar ou descrever os grupos?

-   Por que a aparência de agrupamentos pode ser enganadora?

Algumas dessas questões podem ser respondidas com os dados, enquanto outras irão requerer experiência de domínio sobre os dados.
Muitas delas irão levar você a explorar a relação entre variáveis, por exemplo, para ver se os valores de uma variável podem explicar o comportamento de outra variável.
Chegaremos a isso em breve.

### Valores atípicos

Outliers são observações atípicas, pontos de dados que não estão ajustados ao padrão.
Algumas vezes outliers são erros de entrada de dados.
Algumas vezes são apenas valores extremos que passaram a ser observados nesta coleta de dados, e outras vezes, eles sugerem importantes novas descobertas.
Quando você tem muitos dados, outliers são algumas vezes difíceis de serem vistos em histogramas.
Por exemplo, pegue a distribuição da variável `y` do conjunto de dados de diamante.
A única evidência de valores discrepantes são os limites incomumente amplos no eixo x.

```{r}
#| fig-alt: |
#|   A histogram of lengths of diamonds. The x-axis ranges from 0 to 60 and 
#|   the y-axis ranges from 0 to 12000. There is a peak around 5, and the 
#|   data appear to be completely clustered around the peak.

ggplot(diamante, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
```

Há muitas observações nas caixas comuns que as caixas raras são muito curtas, tornando muito difícil vê-las (embora talvez se você olhar atentamente para 0 você encontre algo).Para facilitar a visualização dos valores incomuns, precisamos ampliar para os valores pequenos do eixo y com `coord_cartesian()`:

```{r}
#| fig-alt: |
#|   A histogram of lengths of diamonds. The x-axis ranges from 0 to 60 and the 
#|   y-axis ranges from 0 to 50. There is a peak around 5, and the data 
#|   appear to be completely clustered around the peak. Other than those data, 
#|   there is one bin at 0 with a height of about 8, one a little over 30 with 
#|   a height of 1 and another one a little below 60 with a height of 1.

ggplot(diamante, aes(x = y)) + 
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

`coord_cartesian()` também possui um argumento `xlim()` para quando você precisar ampliar o eixo x.
ggplot2 também possui as funções `xlim()` e `ylim()` que funcionam de maneira um pouco diferente: elas jogam fora os dados fora dos limites.

Isso nos permite ver que existem três valores incomuns: 0, \~30 e \~60.
Nós retiramos eles com dplyr:

```{r}
#| include: false

old <- options(tibble.print_max = 10, tibble.print_min = 10)
```

```{r}
incomum <- diamante |> 
  filter(y < 3 | y > 20) |> 
  select(preco, x, y, z) |>
  arrange(y)
incomum
```

```{r}
#| include: false

options(old)
```

A variável `y` mede uma das três dimensões desses diamantes, em mm.
Sabemos que os diamantes não podem ter largura de 0 mm, então esses valores devem estar incorretos.
Ao fazer EDA, descobrimos dados faltantes codificados como 0, que nunca teríamos encontrado simplesmente procurando por `NA`s.do a diantes, nós podemos escolher recodificar estes valores como `NA`s, a fim de evitar cálculos enganosos.
Também podemos suspeitar que as medidas de 32 mm e 59 mm são implausíveis: esses diamantes têm mais de uma polegada de comprimento, mas não custam centenas de milhares de dólares!

É uma boa prática repetir suas análises com e sem valores discrepantes.
Se eles tiverem um efeito mínimo nos resultados e você não conseguir descobrir por que estão ali, é razoável omiti-los e seguir em frente.
No entanto, se eles tiverem um efeito substancial nos seus resultados, você não deve abandoná-los sem justificativa.
Você precisará descobrir o que os causou (por exemplo, um erro de entrada de dados) e divulgar que os removeu em seu artigo.

### Exercícios

1.  Explore a distribuição das variáveis `x`, `y`, and `z` em `dados`.
    O que você aprendeu?
    Pense em um diamante e como você pode decidir qual dimensão é o comprimento, largura e profundidade.

2.  Explore a distribuição de `preço`.
    Você percebe algo incomum ou surpreendente?
    (Dica: pense cuidadosamente sobre o `binwidth` e tenha certeza que você tentou uma ampla gama de valores.)

3.  Quantos diamantes equivalem a 0,99 quilates?
    Quantos equivalem a 1 quilate?
    O que você acha que é a causa da diferença?

4.  Compare e contraste `coord_cartesian()` vs. `xlim()` ou `ylim()` quando ampliar em um histograma.
    O que acontece se você deixa o `binwidth` desativado?
    O que acontece se você tentar ampliar de forma que apenas meia barra seja exibida?

## Unusual values {#sec-unusual-values-eda}

If you've encountered unusual values in your dataset, and simply want to move on to the rest of your analysis, you have two options.

1.  Drop the entire row with the strange values:

    ```{r}
    #| eval: false

    diamonds2 <- diamonds |> 
      filter(between(y, 3, 20))
    ```

    We don't recommend this option because one invalid value doesn't imply that all the other values for that observation are also invalid.
    Additionally, if you have low quality data, by the time that you've applied this approach to every variable you might find that you don't have any data left!

2.  Instead, we recommend replacing the unusual values with missing values.
    The easiest way to do this is to use `mutate()` to replace the variable with a modified copy.
    You can use the `if_else()` function to replace unusual values with `NA`:

    ```{r}
    diamonds2 <- diamonds |> 
      mutate(y = if_else(y < 3 | y > 20, NA, y))
    ```

It's not obvious where you should plot missing values, so ggplot2 doesn't include them in the plot, but it does warn that they've been removed:

```{r}
#| dev: "png"
#| fig-alt: |
#|   A scatterplot of widths vs. lengths of diamonds. There is a strong, 
#|   linear association between the two variables. All but one of the diamonds 
#|   has length greater than 3. The one outlier has a length of 0 and a width 
#|   of about 6.5. 

ggplot(diamonds2, aes(x = x, y = y)) + 
  geom_point()
```

To suppress that warning, set `na.rm = TRUE`:

```{r}
#| eval: false

ggplot(diamonds2, aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)
```

Other times you want to understand what makes observations with missing values different to observations with recorded values.
For example, in `nycflights13::flights`[^eda-1], missing values in the `dep_time` variable indicate that the flight was cancelled.
So you might want to compare the scheduled departure times for cancelled and non-cancelled times.
You can do this by making a new variable, using `is.na()` to check if `dep_time` is missing.

[^eda-1]: Remember that when we need to be explicit about where a function (or dataset) comes from, we'll use the special form `package::function()` or `package::dataset`.

```{r}
#| fig-alt: |
#|   A frequency polygon of scheduled departure times of flights. Two lines 
#|   represent flights that are cancelled and not cancelled. The x-axis ranges 
#|   from 0 to 25 minutes and the y-axis ranges from 0 to 10000. The number of 
#|   flights not cancelled are much higher than those cancelled.

nycflights13::flights |> 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + (sched_min / 60)
  ) |> 
  ggplot(aes(x = sched_dep_time)) + 
  geom_freqpoly(aes(color = cancelled), binwidth = 1/4)
```

However this plot isn't great because there are many more non-cancelled flights than cancelled flights.
In the next section we'll explore some techniques for improving this comparison.

### Exercises

1.  What happens to missing values in a histogram?
    What happens to missing values in a bar chart?
    Why is there a difference in how missing values are handled in histograms and bar charts?

2.  What does `na.rm = TRUE` do in `mean()` and `sum()`?

3.  Recreate the frequency plot of `scheduled_dep_time` colored by whether the flight was cancelled or not.
    Also facet by the `cancelled` variable.
    Experiment with different values of the `scales` variable in the faceting function to mitigate the effect of more non-cancelled flights than cancelled flights.

## Covariation

If variation describes the behavior *within* a variable, covariation describes the behavior *between* variables.
**Covariation** is the tendency for the values of two or more variables to vary together in a related way.
The best way to spot covariation is to visualize the relationship between two or more variables.

### A categorical and a numerical variable {#sec-cat-num}

For example, let's explore how the price of a diamond varies with its quality (measured by `cut`) using `geom_freqpoly()`:

```{r}
#| fig-alt: |
#|   A frequency polygon of prices of diamonds where each cut of carat (Fair, 
#|   Good, Very Good, Premium, and Ideal) is represented with a different color 
#|   line. The x-axis ranges from 0 to 30000 and the y-axis ranges from 0 to 
#|   5000. The lines overlap a great deal, suggesting similar frequency 
#|   distributions of prices of diamonds. One notable feature is that 
#|   Ideal diamonds have the highest peak around 1500.

ggplot(diamonds, aes(x = price)) + 
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

Note that ggplot2 uses an ordered color scale for `cut` because it's defined as an ordered factor variable in the data.
You'll learn more about these in @sec-ordered-factors.

The default appearance of `geom_freqpoly()` is not that useful here because the height, determined by the overall count, differs so much across `cut`s, making it hard to see the differences in the shapes of their distributions.

To make the comparison easier we need to swap what is displayed on the y-axis.
Instead of displaying count, we'll display the **density**, which is the count standardized so that the area under each frequency polygon is one.

```{r}
#| fig-alt: |
#|   A frequency polygon of densities of prices of diamonds where each cut of 
#|   carat (Fair, Good, Very Good, Premium, and Ideal) is represented with a 
#|   different color line. The x-axis ranges from 0 to 20000. The lines overlap 
#|   a great deal, suggesting similar density distributions of prices of 
#|   diamonds. One notable feature is that all but Fair diamonds have high peaks 
#|   around a price of 1500 and Fair diamonds have a higher mean than others.

ggplot(diamonds, aes(x = price, y = after_stat(density))) + 
  geom_freqpoly(aes(color = cut), binwidth = 500, linewidth = 0.75)
```

Note that we're mapping the density to `y`, but since `density` is not a variable in the `diamonds` dataset, we need to first calculate it.
We use the `after_stat()` function to do so.

There's something rather surprising about this plot - it appears that fair diamonds (the lowest quality) have the highest average price!
But maybe that's because frequency polygons are a little hard to interpret - there's a lot going on in this plot.

A visually simpler plot for exploring this relationship is using side-by-side boxplots.

```{r}
#| fig-alt: |
#|   Side-by-side boxplots of prices of diamonds by cut. The distribution of 
#|   prices is right skewed for each cut (Fair, Good, Very Good, Premium, and 
#|   Ideal). The medians are close to each other, with the median for Ideal 
#|   diamonds lowest and that for Fair highest.

ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
```

We see much less information about the distribution, but the boxplots are much more compact so we can more easily compare them (and fit more on one plot).
It supports the counter-intuitive finding that better quality diamonds are typically cheaper!
In the exercises, you'll be challenged to figure out why.

`cut` is an ordered factor: fair is worse than good, which is worse than very good and so on.
Many categorical variables don't have such an intrinsic order, so you might want to reorder them to make a more informative display.
One way to do that is with `fct_reorder()`.
You'll learn more about that function in @sec-modifying-factor-order, but we want to give you a quick preview here because it's so useful.
For example, take the `class` variable in the `mpg` dataset.
You might be interested to know how highway mileage varies across classes:

```{r}
#| fig-alt: |
#|   Side-by-side boxplots of highway mileages of cars by class. Classes are 
#|   on the x-axis (2seaters, compact, midsize, minivan, pickup, subcompact, 
#|   and suv).

ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot()
```

To make the trend easier to see, we can reorder `class` based on the median value of `hwy`:

```{r}
#| fig-alt: |
#|   Side-by-side boxplots of highway mileages of cars by class. Classes are 
#|   on the x-axis and ordered by increasing median highway mileage (pickup, 
#|   suv, minivan, 2seater, subcompact, compact, and midsize).

ggplot(mpg, aes(x = fct_reorder(class, hwy, median), y = hwy)) +
  geom_boxplot()
```

If you have long variable names, `geom_boxplot()` will work better if you flip it 90°.
You can do that by exchanging the x and y aesthetic mappings.

```{r}
#| fig-alt: |
#|   Side-by-side boxplots of highway mileages of cars by class. Classes are 
#|   on the y-axis and ordered by increasing median highway mileage.

ggplot(mpg, aes(x = hwy, y = fct_reorder(class, hwy, median))) +
  geom_boxplot()
```

#### Exercises

1.  Use what you've learned to improve the visualization of the departure times of cancelled vs. non-cancelled flights.

2.  Based on EDA, what variable in the diamonds dataset appears to be most important for predicting the price of a diamond?
    How is that variable correlated with cut?
    Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

3.  Instead of exchanging the x and y variables, add `coord_flip()` as a new layer to the vertical boxplot to create a horizontal one.
    How does this compare to exchanging the variables?

4.  One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of "outlying values".
    One approach to remedy this problem is the letter value plot.
    Install the lvplot package, and try using `geom_lv()` to display the distribution of price vs. cut.
    What do you learn?
    How do you interpret the plots?

5.  Create a visualization of diamond prices vs. a categorical variable from the `diamonds` dataset using `geom_violin()`, then a faceted `geom_histogram()`, then a colored `geom_freqpoly()`, and then a colored `geom_density()`.
    Compare and contrast the four plots.
    What are the pros and cons of each method of visualizing the distribution of a numerical variable based on the levels of a categorical variable?

6.  If you have a small dataset, it's sometimes useful to use `geom_jitter()` to avoid overplotting to more easily see the relationship between a continuous and categorical variable.
    The ggbeeswarm package provides a number of methods similar to `geom_jitter()`.
    List them and briefly describe what each one does.

### Two categorical variables

To visualize the covariation between categorical variables, you'll need to count the number of observations for each combination of levels of these categorical variables.
One way to do that is to rely on the built-in `geom_count()`:

```{r}
#| fig-alt: |
#|   A scatterplot of color vs. cut of diamonds. There is one point for each
#|   combination of levels of cut (Fair, Good, Very Good, Premium, and Ideal) 
#|   and color (D, E, F, G, G, I, and J). The sizes of the points represent 
#|   the number of observations for that combination. The legend indicates 
#|   that these sizes range between 1000 and 4000.

ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()
```

The size of each circle in the plot displays how many observations occurred at each combination of values.
Covariation will appear as a strong correlation between specific x values and specific y values.

Another approach for exploring the relationship between these variables is computing the counts with dplyr:

```{r}
diamonds |> 
  count(color, cut)
```

Then visualize with `geom_tile()` and the fill aesthetic:

```{r}
#| fig-alt: |
#|   A tile plot of cut vs. color of diamonds. Each tile represents a 
#|   cut/color combination and tiles are colored according to the number of 
#|   observations in each tile. There are more Ideal diamonds than other cuts, 
#|   with the highest number being Ideal diamonds with color G. Fair diamonds 
#|   and diamonds with color I are the lowest in frequency.

diamonds |> 
  count(color, cut) |>  
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
```

If the categorical variables are unordered, you might want to use the seriation package to simultaneously reorder the rows and columns in order to more clearly reveal interesting patterns.
For larger plots, you might want to try the heatmaply package, which creates interactive plots.

#### Exercises

1.  How could you rescale the count dataset above to more clearly show the distribution of cut within color, or color within cut?

2.  What different data insights do you get with a segmented bar chart if color is mapped to the `x` aesthetic and `cut` is mapped to the `fill` aesthetic?
    Calculate the counts that fall into each of the segments.

3.  Use `geom_tile()` together with dplyr to explore how average flight departure delays vary by destination and month of year.
    What makes the plot difficult to read?
    How could you improve it?

### Two numerical variables

You've already seen one great way to visualize the covariation between two numerical variables: draw a scatterplot with `geom_point()`.
You can see covariation as a pattern in the points.
For example, you can see a positive relationship between the carat size and price of a diamond: diamonds with more carats have a higher price.
The relationship is exponential.

```{r}
#| dev: "png"
#| fig-alt: |
#|   A scatterplot of price vs. carat. The relationship is positive, somewhat 
#|   strong, and exponential.

ggplot(smaller, aes(x = carat, y = price)) +
  geom_point()
```

(In this section we'll use the `smaller` dataset to stay focused on the bulk of the diamonds that are smaller than 3 carats)

Scatterplots become less useful as the size of your dataset grows, because points begin to overplot, and pile up into areas of uniform black, making it hard to judge differences in the density of the data across the 2-dimensional space as well as making it hard to spot the trend.
You've already seen one way to fix the problem: using the `alpha` aesthetic to add transparency.

```{r}
#| dev: "png"
#| fig-alt: |
#|   A scatterplot of price vs. carat. The relationship is positive, somewhat 
#|   strong, and exponential. The points are transparent, showing clusters where 
#|   the number of points is higher than other areas, The most obvious clusters 
#|   are for diamonds with 1, 1.5, and 2 carats.

ggplot(smaller, aes(x = carat, y = price)) + 
  geom_point(alpha = 1 / 100)
```

But using transparency can be challenging for very large datasets.
Another solution is to use bin.
Previously you used `geom_histogram()` and `geom_freqpoly()` to bin in one dimension.
Now you'll learn how to use `geom_bin2d()` and `geom_hex()` to bin in two dimensions.

`geom_bin2d()` and `geom_hex()` divide the coordinate plane into 2d bins and then use a fill color to display how many points fall into each bin.
`geom_bin2d()` creates rectangular bins.
`geom_hex()` creates hexagonal bins.
You will need to install the hexbin package to use `geom_hex()`.

```{r}
#| layout-ncol: 2
#| fig-width: 3
#| fig-alt: |
#|   Plot 1: A binned density plot of price vs. carat. Plot 2: A hexagonal bin 
#|   plot of price vs. carat. Both plots show that the highest density of 
#|   diamonds have low carats and low prices.

ggplot(smaller, aes(x = carat, y = price)) +
  geom_bin2d()

# install.packages("hexbin")
ggplot(smaller, aes(x = carat, y = price)) +
  geom_hex()
```

Another option is to bin one continuous variable so it acts like a categorical variable.
Then you can use one of the techniques for visualizing the combination of a categorical and a continuous variable that you learned about.
For example, you could bin `carat` and then for each group, display a boxplot:

```{r}
#| fig-alt: |
#|   Side-by-side box plots of price by carat. Each box plot represents diamonds 
#|   that are 0.1 carats apart in weight. The box plots show that as carat 
#|   increases the median price increases as well. Additionally, diamonds with 
#|   1.5 carats or lower have right skewed price distributions, 1.5 to 2 have 
#|   roughly symmetric price distributions, and diamonds that weigh more have 
#|   left skewed distributions. Cheaper, smaller diamonds have outliers on the 
#|   higher end, more expensive, bigger diamonds have outliers on the lower end.

ggplot(smaller, aes(x = carat, y = price)) + 
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

`cut_width(x, width)`, as used above, divides `x` into bins of width `width`.
By default, boxplots look roughly the same (apart from number of outliers) regardless of how many observations there are, so it's difficult to tell that each boxplot summaries a different number of points.
One way to show that is to make the width of the boxplot proportional to the number of points with `varwidth = TRUE`.

#### Exercises

1.  Instead of summarizing the conditional distribution with a boxplot, you could use a frequency polygon.
    What do you need to consider when using `cut_width()` vs. `cut_number()`?
    How does that impact a visualization of the 2d distribution of `carat` and `price`?

2.  Visualize the distribution of `carat`, partitioned by `price`.

3.  How does the price distribution of very large diamonds compare to small diamonds?
    Is it as you expect, or does it surprise you?

4.  Combine two of the techniques you've learned to visualize the combined distribution of cut, carat, and price.

5.  Two dimensional plots reveal outliers that are not visible in one dimensional plots.
    For example, some points in the following plot have an unusual combination of `x` and `y` values, which makes the points outliers even though their `x` and `y` values appear normal when examined separately.
    Why is a scatterplot a better display than a binned plot for this case?

    ```{r}
    #| eval: false
    diamonds |> 
      filter(x >= 4) |> 
      ggplot(aes(x = x, y = y)) +
      geom_point() +
      coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
    ```

6.  Instead of creating boxes of equal width with `cut_width()`, we could create boxes that contain roughly equal number of points with `cut_number()`.
    What are the advantages and disadvantages of this approach?

    ```{r}
    #| eval: false
    ggplot(smaller, aes(x = carat, y = price)) + 
      geom_boxplot(aes(group = cut_number(carat, 20)))
    ```

## Patterns and models

If a systematic relationship exists between two variables it will appear as a pattern in the data.
If you spot a pattern, ask yourself:

-   Could this pattern be due to coincidence (i.e. random chance)?

-   How can you describe the relationship implied by the pattern?

-   How strong is the relationship implied by the pattern?

-   What other variables might affect the relationship?

-   Does the relationship change if you look at individual subgroups of the data?

Patterns in your data provide clues about relationships, i.e., they reveal covariation.
If you think of variation as a phenomenon that creates uncertainty, covariation is a phenomenon that reduces it.
If two variables covary, you can use the values of one variable to make better predictions about the values of the second.
If the covariation is due to a causal relationship (a special case), then you can use the value of one variable to control the value of the second.

Models are a tool for extracting patterns out of data.
For example, consider the diamonds data.
It's hard to understand the relationship between cut and price, because cut and carat, and carat and price are tightly related.
It's possible to use a model to remove the very strong relationship between price and carat so we can explore the subtleties that remain.
The following code fits a model that predicts `price` from `carat` and then computes the residuals (the difference between the predicted value and the actual value).
The residuals give us a view of the price of the diamond, once the effect of carat has been removed.
Note that instead of using the raw values of `price` and `carat`, we log transform them first, and fit a model to the log-transformed values.
Then, we exponentiate the residuals to put them back in the scale of raw prices.

```{r}
#| message: false
#| dev: "png"
#| fig-alt: |
#|   A scatterplot of residuals vs. carat of diamonds. The x-axis ranges from 0 
#|   to 5, the y-axis ranges from 0 to almost 4. Much of the data are clustered 
#|   around low values of carat and residuals. There is a clear, curved pattern 
#|   showing decrease in residuals as carat increases.

library(tidymodels)

diamonds <- diamonds |>
  mutate(
    log_price = log(price),
    log_carat = log(carat)
  )

diamonds_fit <- linear_reg() |>
  fit(log_price ~ log_carat, data = diamonds)

diamonds_aug <- augment(diamonds_fit, new_data = diamonds) |>
  mutate(.resid = exp(.resid))

ggplot(diamonds_aug, aes(x = carat, y = .resid)) + 
  geom_point()
```

Once you've removed the strong relationship between carat and price, you can see what you expect in the relationship between cut and price: relative to their size, better quality diamonds are more expensive.

```{r}
#| fig-alt: |
#|   Side-by-side box plots of residuals by cut. The x-axis displays the various 
#|   cuts (Fair to Ideal), the y-axis ranges from 0 to almost 5. The medians are 
#|   quite similar, between roughly 0.75 to 1.25. Each of the distributions of 
#|   residuals is right skewed, with many outliers on the higher end.

ggplot(diamonds_aug, aes(x = cut, y = .resid)) + 
  geom_boxplot()
```

We're not discussing modelling in this book because understanding what models are and how they work is easiest once you have tools of data wrangling and programming in hand.

## Summary

In this chapter you've learned a variety of tools to help you understand the variation within your data.
You've seen techniques that work with a single variable at a time and with a pair of variables.
This might seem painfully restrictive if you have tens or hundreds of variables in your data, but they're the foundation upon which all other techniques are built.

In the next chapter, we'll focus on the tools we can use to communicate our results.
